---
title: "Lake Lugano brGDGT vs 16S-rDNA"
output: 
   html_document: 
      
      code_folding: show
      standalone: true
      self-contained: true
   pdf_document: default
   html_notebook: default
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 include = TRUE,
 results = "show",
 cache = F,
 warning = FALSE,
 message = FALSE,
 fig.path = "figs/"
 # dev = c("pdf", "png")
 #dev.args = list(pdf = list(encoding = "WinAnsi", useDingbats = FALSE))
 )

require(tinytex)
```
&nbsp;

#### Reproducing the analysis published in: *Yuki Weber, Jaap S. Sinninghe Damsté, Jakob Zopfi, Cindy De Jonge, Adrian Gili, Carsten J. Schubert, Fabio Lepori, Moritz F. Lehmann, Helge Niemann (2018): **"Redox-dependent niche differentiation of tetraether lipid producing bacteria: Evidence for multiple branched GDGT sources in lakes"***

&nbsp;

### Preparation

1.) Install 'R' and 'RStudio' on your machine.

2.) Download Supplementary Information Datasets 1 and 2 from the publisher's website and place files *dataset1.xls* and *dataset2.xls* into the working directory (i.e, the RStudio project folder).  

*<b>note:</b>*
To execute the following code in RStudio, run the respective code chunks (gray boxes) either by pressing *CMD + Shift + Enter* (Mac), or by clicking on the 'play' button in the upper right corner of rach code block.

&nbsp;

### Run the analysis

&nbsp;

#### Install phylo.lipids package
```{r}
# install devtools if not present and load
if (!"devtools" %in% rownames(installed.packages())) {
 install.packages("devtools")
}
require(devtools)

# install the phylo.lipids package from local source and load
if (!"phylo.lipids" %in% rownames(installed.packages())) {
 devtools::install_github("yukiweber/phylo.lipids", build_vignettes = F)
}
require(phylo.lipids)
```

&nbsp;

#### Load raw data files
```{r, echo=TRUE, include=TRUE, warning=T}
### load data files ----
## OTU table
lugano_otu =
  readxl::read_xls ("dataset1.xls", sheet = "otu.table" ) %>%
  as.data.frame()
# set otu names as rownames 
rownames(lugano_otu) = lugano_otu[, 1][] 
# remove name column
lugano_otu = lugano_otu[ ,-1]
  
## total DNA concentrations
lugano_sam_dat =
  readxl::read_xls ("dataset1.xls", sheet = "sample.data" ) %>%
  as.data.frame()
# set rownames as sample names (identical with otu table)
rownames(lugano_sam_dat) = colnames(lugano_otu)

## taxonomy table
lugano_tax =
  readxl::read_xls ("dataset1.xls", sheet = "tax.table" ) %>%
  as.data.frame()
# set otu names as rownames 
rownames(lugano_tax) = lugano_tax[,1][]
# remove name column
lugano_tax = lugano_tax[ ,-1]

## build phyloseq object ----
## contains OTU and taxonmoy tables, and sample data (here: DNA concentraions)
lugano_PHY =
  phyloseq::phyloseq (
    phyloseq::otu_table (as.matrix(lugano_otu), taxa_are_rows = TRUE),
    phyloseq::tax_table (as.matrix(lugano_tax)),
    phyloseq::sample_data (lugano_sam_dat)
  )
```

&nbsp;

#### Processing of DNA data
```{r, echo=TRUE, include=TRUE, warning=FALSE}
## Transform OTU data ----
lugano_PHY %>%
# remove taxa that have less than 40 reads across all samples
phylo.lipids::trim_taxa_sum (min = 40) %>% 
# remove taxa that occurr in less than 5 samples
phylo.lipids::trim_taxa_samp (min = 5, fraction = F) %>%
# normalize to total dna
phylo.lipids::norm_dna (.@sam_data$`yg.DNA.L-1`) %>%
# remove eucaryotes and archaea
phylo.lipids::remove_clades (., c("Chloroplast", "Mitochondria", "Archaea")) -> PHY1

PHY2 = 
  PHY1 %>%
  # remove 100 m sample because the corresponsding lipid sample was lost
  phyloseq::subset_samples (depth != 100)

# test if there are still eukaryotic sequences
if (class (try (phyloseq::subset_taxa(PHY2, Class=="Chloroplast"))) != "try-error" | 
    class (try (phyloseq::subset_taxa(PHY2, Family=="Mitochondria"))) != "try-error"
) { rm(PHY2)
    rep("ALARM: still containing Eukaryotes!!",6) }

## differentiate between classes of Proteobacteria ----
PHY2 %>%
 phylo.lipids::diff_proteo() -> PHY3
```

&nbsp;

#### Load and prepare raw lipid data
```{r, echo=TRUE, include=TRUE, warning=TRUE}
## load lipid data
lugano_SPM =
  readxl::read_xls ("dataset2.xls", sheet = "Ludano.SPM.ng.L-1" ) %>%
  as.data.frame()
## replace introduced dots in lipid names
colnames(lugano_SPM) = gsub("[.]","'", colnames(lugano_SPM))

## select lipid fraction for correlation (here: sum of all fractions, i.e., "total")
lipids =
 lugano_SPM %>% dplyr::filter (fraction == "total") %>%
 ## and keep only lipid data columns 
 dplyr::select (-depth, - fraction)

# subset lipids ----
# (here we only look at the major brGDGTs and the sum of all brGDGTs)
# the brGDGT names (columns)
bG.major = c("IIIa","IIIa''","IIIa'","IIa", "IIa'","Ia")
# select major brGDGTs and sum
lipids %>% dplyr::select(bG.major, sum_bG) -> L
```

&nbsp;

#### This yields a class phyloseq object and a lipid table
```{r}
# taxonomy and OTU-specific DNA concentrations:
# (nuber of samples must match number of rows in the lipid table)
PHY3
# lipid concentrations:
# (make sure all colums contain lipid data, i.e., no name/ID column etc)
tibble::as.tibble(lipids) %>% knitr::kable(digits = 2)
```

&nbsp;

#### Correlation between OTUs and each lipid
```{r lugano_cor_phy, fig.height=4}
# use the cor_phy() function and save the results
# for information on the function's options, refer to the help page
phylo.lipids::cor_phy( phy= PHY3, L,
                       display = c("Subgroup 21"),
                       alpha = 0.9,
                       alpha.other = 0.2,
                       col.other = "gray70",
                       colors = "random",
                       luminosity = "dark",
                       width = 0.3,
                       aspect = 1,
                       box.plot = F
                       ) -> lugano_cor_phy_results

# display output
# output contains a plot and two data frames
lugano_cor_phy_results$plot
lugano_cor_phy_results$lipids 
lugano_cor_phy_results$taxa 

PHY3 %>% 
   #trim_taxa_sum(10) %>%
   #tax_glom("Order") %>% 
   phyloseq::plot_net(type ="samples",
                      point_alpha= 0.8,
                      hjust = -1,
                      point_label = "depth",
                      color="depth",
                      maxdist = 0.75) -> net
net
PHY3@sam_data

PHY3 %>% 
   trim_taxa_samp(14) %>%
   tax_glom("Phylum") %>% 
   phyloseq::plot_net(type ="taxa",
                      point_alpha= 0.8,
                      hjust = -0.2,
                      point_label = "Phylum",
                      color="Phylum",
                      maxdist = 0.75) -> net
net
PHY3@sam_data



```

&nbsp;

#### Diversity of OTUs correlated with each lipid
```{r lugano_cor_tax, include=T, results='show', echo=T, fig.width=8, fig.asp=1, fig.height=8}

# Subset major brGDGTs
L = lipids  %>% dplyr::select(bG.major)

# use phylo_lipids() function to perform pearson correlation and
# summarize taxonomy at the Phylum level 
phylo_lipids( phy = PHY3, L,
              level = "Phylum",
              thresh = 0.75,
              method = "pearson",
              cutoff = 0.04,
              exempt = c("Acidobacteria"),
              n.col = 3,
              sort.tax = "alpha",
              text.size = 14,
              guide.col = 1
              ) -> lugano_cor_tax_results


lugano_cor_tax_results$bars + ggplot2::theme(aspect.ratio = 0.5)
lugano_cor_tax_results$pies
lugano_cor_tax_results$table
lugano_cor_tax_results$summary
```


#### Diversity of Acidobacteria correlated with each lipid
```{r lugano_acido_cor_tax, fig.width=8, fig.asp=1, fig.height=8}

PHY3 %>%
   phyloseq::subset_taxa(Phylum=="Acidobacteria") %>%
   phylo_lipids( phy = ., 
                 lipids = L,
                 level = "Class",
                 thresh = 0.75,
                 method = "pearson",
                 cutoff = 0.0,
                 n.col = 3,
                 sort.tax = "alpha",
                 text.size = 14
                 ) -> lugano_acido_cor_tax_results

lugano_acido_cor_tax_results$bars + ggplot2::theme(aspect.ratio = 0.5)
lugano_acido_cor_tax_results$pies
lugano_acido_cor_tax_results$table
lugano_acido_cor_tax_results$summary
```

