---
title: "Lake Lugano 16S-rDNA ecology"
author: "Yuki Weber"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 include = TRUE,
 results = "show",
 cache = F,
 warning = FALSE,
 message = FALSE,
 fig.path = "figs/"
 # dev = c("pdf", "png")
 #dev.args = list(pdf = list(encoding = "WinAnsi", useDingbats = FALSE))
 )

#require(tinytex)
```
&nbsp;
## Exploring bacterial diversity in ecological niches within the water column

&nbsp;
#### Install phylo.lipids package
```{r}
# install devtools if not present and load
if (!"devtools" %in% rownames(installed.packages())) {
 install.packages("devtools")
}
require(devtools)

# install the phylo.lipids development package and load 
if (!"phylo.lipids" %in% rownames(installed.packages())) {
 devtools::install_github("yukiweber/phylo.lipids", ref = "release/0.0.901", build_vignettes = T)
}
require(phylo.lipids)
```

&nbsp;

#### Load raw data files
(alternatively you can use your existing phyloseq objects)  
Be aware of the effects of using either relative abundance data or DNA concentrations 
(based on total DNA or qPCR etc.)

```{r, echo=TRUE, include=TRUE, warning=T}
### load data files ----
otu_names=
  readr::read_csv(
    system.file("extdata", "lugano.otu.table.csv", package = "phylo.lipids"),
    col_names = T ) %>% 
  as.data.frame() %>%
  dplyr::select(1) %>%
  as.vector() 
otu_names = otu_names[,1] 

otu =
  readr::read_csv(
    system.file("extdata", "lugano.otu.table.csv", package = "phylo.lipids"),
    col_names = T ) %>%
    as.data.frame() %>%
    dplyr::select(-1)
rownames(otu) = otu_names

sam_dat=
  readr::read_csv(
    system.file("extdata", "lugano.sample.data.csv", package = "phylo.lipids"),
    col_names = T ) %>% 
    as.data.frame()
rownames(sam_dat) = sam_dat$sample

tax = 
  readr::read_csv(
    system.file("extdata", "lugano.tax.table.csv", package = "phylo.lipids"),
    col_names = T ) %>%
    as.data.frame() %>%
    dplyr::select(-1)
rownames(tax) = otu_names

## build a phyloseq object ----
PHY= phyloseq::phyloseq (phyloseq::otu_table (as.matrix(otu), taxa_are_rows = TRUE),
                         phyloseq::tax_table (as.matrix(tax)),
                         phyloseq::sample_data (sam_dat)
                         )

#### Processing of DNA data

## Transform OTU data ----
PHY %>%
# remove taxa that have less than 40 reads across all samples
phylo.lipids::trim_taxa_sum (min = 40) %>% 
# remove taxa that occurr in less than 5 samples
phylo.lipids::trim_taxa_samp (min = 5, fraction = F) %>%
# normalize to total dna
phylo.lipids::norm_dna (dna = sam_dat$`µg.DNA.L-1`) %>%
# remove eucaryotes and archaea
phylo.lipids::remove_clades (., c("Chloroplast", "Mitochondria", "Archaea")) -> PHY1

PHY2 = 
  PHY1 %>%
  # remove 100 m sample because the corresponsding lipid sample was lost
  phyloseq::subset_samples (depth != 100)

# test if there are still eukaryotic sequences
if (class (try (phyloseq::subset_taxa(PHY2, Class=="Chloroplast"))) != "try-error" | 
    class (try (phyloseq::subset_taxa(PHY2, Family=="Mitochondria"))) != "try-error"
) { rm(PHY2)
    rep("ALARM: still containing Eukaryotes!!",6) }

## differentiate between classes of Proteobacteria ----
PHY2 %>%
 phylo.lipids::diff_proteo() -> PHY3
```

&nbsp;

#### First explore the ecological niche distribution
```{r, echo=TRUE, include=TRUE, warning=TRUE}

## try to find 'k' values that give a good separation of the groups 
## pay attention to the red ribbons (2 SD error bounds)
## eventuallt set 'otus = TRUE'
## use the depth label to reverse scales and mek them numeric
## 'depth' has to be a column of phy@sam_data
## use 'count = "sum"' to use the actual reads/concentrations/relative abundance values
## if 'count = "n"', the NUMBER OF OTUS affiliated with each taxonomic group (e.g. Phylum) will be used,
## regardless of the abundance of the indivisual otus
 

plot_groups(phy = PHY3,
            label = "Depth",
            k=22,
            otus = T,
            level = "Phylum",
            fancy = F,
            count = "sum",
            ncol = 10
            ) -> GG0

GG0$facet_groups

```

Then look at the phylogenetic diversity in each group. 

```{r}
GG0$facet_pies
```

Eventually adjust the cutoff value to reduce the number of taxonomic units.  
You exempt taxa of interest from the cutoff (i.e., always show them) and/or highlight taxa of interest.

```{r}

plot_groups(phy = PHY3,
            label = "depth",
            k=22,
            otus = T,
            level = "Phylum",
            fancy = F,
            count = "sum",
            cutoff = 0.15, ################### 15% cutoff
            exempt = "Nitrospirae"
            ) -> GG0

GG0$facet_pies

plot_groups(phy = PHY3,
            label = "depth",
            k=22,
            otus = T,
            level = "Phylum",
            fancy = F,
            count = "sum",
            cutoff = 0.15, # 15% cutoff
            exempt = c("Nitrospirae","Acidobacteria"), ############ !
            highlight = c("Nitrospirae","Acidobacteria") ############# !
            ) -> GG0

GG0$facet_pies


```

&nbsp;

Now let's have a closer look at the phylogenetic diversity in SELECTED cluster groups of interest

```{r}

## set fancy = TRUE to make overlay plots
## It's slow so do not use it on the full data set

plot_groups(phy = PHY3,
            label = "depth",
            k=22,
            otus = T,
            level = "Family", ######################### !
            fancy = T, ############################### !
            count = "sum",
            cutoff = 0.05, 
            exempt = c("Methylococcaceae","Crenotrichaceae"),
            prune_groups = c("G3","G5","G21", "G19", "G14"), ### !!!
            key_col = 2
            ) -> GG1

plot_groups(phy = PHY3,
            label = "depth",
            k=22,
            otus = T,
            level = "Family",
            fancy = T,
            count = "sum",
            cutoff = 0.05, 
            exempt = c("Methylococcaceae","Crenotrichaceae","Nitrospirales Incertae Sedis"),
            prune_groups = c("G3","G5","G21", "G19", "G14"),
            highlight = c("Methylococcaceae","Crenotrichaceae","Nitrospirales Incertae Sedis"),
            key_col = 2
            ) -> GG2
```
```{r}
GG1$groups_pies_full
GG2$groups_pies_full
```



